---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/0.1.0/active.svg)](http://www.repostatus.org/#active)
 [![Build Status](https://travis-ci.org/jennybc/googlesheets.svg?branch=master)](https://travis-ci.org/jennybc/googlesheets) [![Coverage Status](https://coveralls.io/repos/jennybc/googlesheets/badge.svg)](https://coveralls.io/r/jennybc/googlesheets)

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r make-clean, echo = FALSE, include = FALSE}
## if previous compilation errored out, intended clean up may be incomplete
suppressWarnings(
  file.remove(c("~/tmp/gapminder-africa.csv", "~/tmp/gapminder.xlsx")))
googlesheets::gs_vecdel(c("foo", "mini-gap", "iris"), verbose = FALSE)
```
---

Google Sheets R API
---

Access and manage Google spreadsheets from R with `googlesheets`. 

Features:

  - Access a spreadsheet by its title, key or URL.
  - Extract data or edit data.
  - Create | delete | rename | copy | upload | download spreadsheets and worksheets.

`googlesheets` is inspired by [gspread](https://github.com/burnash/gspread), a Google Spreadsheets Python API

The exuberant prose in this README is inspired by [Tabletop.js](https://github.com/jsoma/tabletop): If you've ever wanted to get data in or out of a Google Spreadsheet from R without jumping through a thousand hoops, welcome home!

#### What the hell do I do with this?

Think of `googlesheets` as a read/write CMS that you (or your less R-obsessed friends) can edit through Google Docs, as well via R. It's like Christmas up in here.

Use a [Google Form](http://www.google.com/forms/about/) to conduct a survey, which populates a Google Sheet.

Gather data while you're in the field in a Google Sheet, maybe [with an iPhone](https://itunes.apple.com/us/app/google-sheets/id842849113?mt=8) or [an Android device](https://play.google.com/store/apps/details?id=com.google.android.apps.docs.editors.sheets&hl=en). Take advantage of [data validation](https://support.google.com/docs/answer/139705?hl=en) to limit the crazy on the way in.

There are various ways to harvest web data directly into a Google Sheet. For example:

  * [This blog post](http://blog.aylien.com/post/114757623598/sentiment-analysis-of-restaurant-reviews) from Aylien.com has a simple example that uses the `=IMPORTXML()` formula to populate a Google Sheet with restaurant reviews and ratings from TripAdvisor.
  * Martin Hawksey offers [TAGS](https://tags.hawksey.info), a free Google Sheet template to setup and run automated collection of search results from Twitter.
  * Martin Hawksey also has a great blog post, [Feeding Google Spreadsheets](https://mashe.hawksey.info/2012/10/feeding-google-spreadsheets-exercises-in-import/), that demonstrates how functions like `importHTML`, `importFeed`, and `importXML` help you get data from the web into a Google Sheet with no programming.
  * Martin Hawksey has another blog post about [feeding a Google Sheet from IFTTT](https://mashe.hawksey.info/2012/09/ifttt-if-i-do-that-on-insert-social-networkrss-feedother-then-add-row-to-google-spreadsheet/). [IFTTT](https://ifttt.com) stands for "if this, then that" and it's "a web-based service that allows users to create chains of simple conditional statements, called 'recipes', which are triggered based on changes to other web services such as Gmail, Facebook, Instagram, and Craigslist" (from [Wikipedia](http://en.wikipedia.org/wiki/IFTTT)).

Use `googlesheets` to get all that data into R.

Use it in a Shiny app! *this will be the next demo/vignette I write*

What other ideas do you have?
  
### Install googlesheets

```{r eval = FALSE}
devtools::install_github("jennybc/googlesheets")
```

*We plan to submit to CRAN in June 2015, so feedback on functionality and usability is especially valuable to us now!*

### Take a look at the vignette

No, actually, __don't__. This README is much more current than the vignette, though that will have to change soon!

If you insist, [check out the current state of the vignette](http://htmlpreview.github.io/?https://raw.githubusercontent.com/jennybc/googlesheets/master/vignettes/basic-usage.html).

### Here are some draft slides

[Draft slides](https://speakerdeck.com/jennybc/googlesheets-draft) for a talk in July 2015

### Load googlesheets

`googlesheets` is designed for use with the `%>%` pipe operator and, to a lesser extent, the data-wrangling mentality of [`dplyr`](http://cran.r-project.org/web/packages/dplyr/index.html). This README uses both, but the examples in the help files emphasize usage with plain vanilla R, if that's how you roll. `googlesheets` uses `dplyr` internally but does not require the user to do so. You can make the `%>%` pipe operator available in your own work by loading [`dplyr`](http://cran.r-project.org/web/packages/dplyr/index.html) or [`magrittr`](http://cran.r-project.org/web/packages/magrittr/index.html).

```{r load-package}
library("googlesheets")
suppressPackageStartupMessages(library("dplyr"))
```

### Function naming convention

All functions start with `gs_`, which plays nicely with tab completion in RStudio, for example. If the function has something to do with worksheets or tabs within a spreadsheet, then it will start with `gs_ws_`.

### See some spreadsheets you can access

The `gs_ls()` function returns the sheets you would see in your Google Sheets home screen: <https://docs.google.com/spreadsheets/>. This should include sheets that you own and may also show sheets owned by others but that you are permitted to access, if you visited the sheet in the browser. Expect a prompt to authenticate yourself in the browser at this point (more below re: authentication).

```{r list-sheets}
(my_sheets <- gs_ls())
# (expect a prompt to authenticate with Google interactively HERE)
my_sheets %>% glimpse()
```

### Get a Google spreadsheet to practice with

Don't worry if you don't have any suitable Google Sheets lying around! We've published a sheet for you to practice with and have built functions into `googlesheets` to help you access it. The example sheet holds some of the [Gapminder data](https://github.com/jennybc/gapminder); feel free to [visit the Sheet in the browser](https://w3id.org/people/jennybc/googlesheets_gap_url). The code below will put a copy of this sheet into your Drive, titled "Gapminder".

```{r copy-gapminder, eval = FALSE}
gs_gap() %>% 
  gs_copy(to = "Gapminder")
```

If that seems to have worked, go check for a sheet named "Gapminder" in your Google Sheets home screen: <https://docs.google.com/spreadsheets/>. You could also run `gs_ls()` again and make sure the Gapminder sheet is listed.

### Register a spreadsheet

If you plan to consume data from a sheet or edit it, you must first __register__ it. This is how `googlesheets` records important info about the sheet that is required downstream by the Google Sheets or Google Drive APIs. Once registered, you can print the result to get some basic info about the sheet.

`googlesheets` provides several registration functions. Specifying the sheet by title? Use `gs_title()`. By key? Use `gs_key()`. You get the idea.

*We're using the built-in functions `gs_gap_key()` and `gs_gap_url()` to produce the key and browser URL for the Gapminder example sheet, so you can see how this will play out with your own projects.*

```{r register-sheet}
gap <- gs_title("Gapminder")
gap

# Need to access a sheet you do not own?
# Access it by key if you know it!
(GAP_KEY <- gs_gap_key())
third_party_gap <- GAP_KEY %>%
  gs_key()

# Need to access a sheet you do not own but you have a sharing link?
# Access it by URL!
(GAP_URL <- gs_gap_url())
third_party_gap <- GAP_URL %>%
  gs_url()
# note: registration via URL may not work for "old" sheets

# Worried that a spreadsheet's registration is out-of-date?
# Re-register it!
gap <- gap %>% gs_gs()
```

The registration functions `gs_title()`, `gs_key()`, `gs_url()`, and `gs_gs()` return a registered sheet as a `googlesheet` object, which is the first argument to practically every function in this package. Likewise, almost every function returns a freshly registered `googlesheet` object, ready to be stored or piped into the next command.

*We export a utility function, `extract_key_from_url()`, to help you get and store the key from a browser URL. Registering via browser URL is fine, but registering by key is probably a better idea in the long-run.*

### Consume data

#### Ignorance is bliss

If you want to consume the data in a worksheet and get something rectangular back, use the all-purpose function `gs_read()`. By default, it reads all the data in a worksheet.

```{r}
oceania <- gap %>% gs_read(ws = "Oceania")
oceania
str(oceania)
glimpse(oceania)
```

You can target specific cells via the `range =` argument. The simplest usage is to specify an Excel-like cell range, such as range = "D12:F15" or range = "R1C12:R6C15". The cell rectangle can be specified in various other ways, using helper functions.

```{r}
gap %>% gs_read(ws = 2, range = "A1:D8")
gap %>% gs_read(ws = "Europe", range = cell_rows(1:4))
gap %>% gs_read(ws = "Europe", range = cell_rows(100:103), col_names = FALSE)
gap %>% gs_read(ws = "Africa", range = cell_cols(1:4))
gap %>% gs_read(ws = "Asia", range = cell_limits(c(1, 4), c(5, NA)))
```

`gs_read()` is a wrapper that bundles together the most common methods to read data from the API and transform it for downstream use. You can refine it's behavior further, by passing more arguments via `...`. Read the help file for more details.

If `gs_read()` doesn't do what you need, then keep reading for the underlying functions to read and post-process data.

#### Specify the consumption method

There are three ways to consume data from a worksheet within a Google spreadsheet. The order goes from fastest-but-more-limited to slowest-but-most-flexible:

  * `gs_read_csv()`: Don't let the name scare you! Nothing is written to file during this process. The name just reflects that, under the hood, we request the data via the "exportcsv" link. For cases where `gs_read_csv()` and `gs_read_listfeed()` both work, we see that `gs_read_csv()` is around __50 times faster__. Use this when your data occupies a nice rectangle in the sheet and you're willing to consume all of it. You will get a `tbl_df` back, which is basically just a `data.frame`. In fact, you might want to use `gs_read_csv()`, it in other, less tidy scenarios and do further munging in R.
  * `gs_read_listfeed()`: Gets data via the ["list feed"](https://developers.google.com/google-apps/spreadsheets/#working_with_list-based_feeds), which consumes data row-by-row. Like `gs_read_csv()`, this is appropriate when your data occupies a nice rectangle. You will again get a `tbl_df` back, but your variable names may have been mangled (by Google, not us!). Specifically, variable names will be forcefully lowercased and all non-alpha-numeric characters will be removed. Why do we even have this function? The list feed supports some query parameters for sorting and filtering the data, which we plan to support (#17).
  * `gs_read_cellfeed()`: Get data via the ["cell feed"](https://developers.google.com/google-apps/spreadsheets/#working_with_cell-based_feeds), which consumes data cell-by-cell. This is appropriate when you want to consume arbitrary cells, rows, columns, and regions of the sheet. It is invoked by `gs_read()` whenever the `range =` argument is used. It works great for modest amounts of data but can be rather slow otherwise. `gs_read_cellfeed()` returns a `tbl_df` with __one row per cell__. You can target specific cells via the `range` argument. See below for demos of `gs_reshape_cellfeed()` and `gs_simplify_cellfeed()` which help with post-processing.

```{r csv-list-and-cell-feed}
# Get the data for worksheet "Oceania": the super-fast csv way
oceania_csv <- gap %>% gs_read_csv(ws = "Oceania")
str(oceania_csv)
oceania_csv

# Get the data for worksheet "Oceania": the less-fast tabular way ("list feed")
oceania_list_feed <- gap %>% gs_read_listfeed(ws = "Oceania") 
str(oceania_list_feed)
oceania_list_feed

# Get the data for worksheet "Oceania": the slow cell-by-cell way ("cell feed")
oceania_cell_feed <- gap %>% gs_read_cellfeed(ws = "Oceania") 
str(oceania_cell_feed)
oceania_cell_feed
```

#### Quick speed comparison

Let's consume all the data for Africa by all 3 methods and see how long it takes.

```{r}
jfun <- function(readfun)
  system.time(do.call(readfun, list(gs_gap(), ws = "Africa", verbose = FALSE)))
readfuns <- c("gs_read_csv", "gs_read_listfeed", "gs_read_cellfeed")
readfuns <- sapply(readfuns, get, USE.NAMES = TRUE)
sapply(readfuns, jfun)
```

#### Post-processing data from the cell feed

If you consume data from the cell feed with `gs_read_cellfeed(..., range = ...)`, you get a data.frame back with **one row per cell**. The package offers two functions to post-process this into something more useful, `gs_reshape_cellfeed()` and `gs_simplify_cellfeed()`.

To reshape into a table, use `gs_reshape_cellfeed()`. You can signal that the first row contains column names (or not) with `col_names = TRUE` (or `FALSE`). Or you can provide a character vector of names. This is inspired by the `col_names` argument of `readxl::read_excel()` and `readr::read_delim()`, which generalizes the `header` argument of `read.table()`.

```{r post-processing}
# Reshape: instead of one row per cell, make a nice rectangular data.frame
australia_cell_feed <- gap %>%
  gs_read_cellfeed(ws = "Oceania", range = "A1:F13") 
str(australia_cell_feed)
oceania_cell_feed
australia_reshaped <- australia_cell_feed %>% gs_reshape_cellfeed()
str(australia_reshaped)
australia_reshaped

# Example: first 3 rows
gap_3rows <- gap %>% gs_read_cellfeed("Europe", range = cell_rows(1:3))
gap_3rows %>% head()

# convert to a data.frame (by default, column names found in first row)
gap_3rows %>% gs_reshape_cellfeed()

# arbitrary cell range, column names no longer available in first row
gap %>%
  gs_read_cellfeed("Oceania", range = "D12:F15") %>%
  gs_reshape_cellfeed(col_names = FALSE)

# arbitrary cell range, direct specification of column names
gap %>%
  gs_read_cellfeed("Oceania", range = cell_limits(c(2, 1), c(5, 3))) %>%
  gs_reshape_cellfeed(col_names = paste("thing", c("one", "two", "three"),
                                        sep = "_"))
```

To extract the cell data into an atomic vector, possibly named, use `gs_simplify_cellfeed()`. You can signal that the first row contains column names (or not) with `col_names = TRUE` (or `FALSE`). There are several arguments to control conversion.

```{r}
# Example: first row only
gap_1row <- gap %>% gs_read_cellfeed("Europe", range = cell_rows(1))
gap_1row

# convert to a named character vector
gap_1row %>% gs_simplify_cellfeed()

# Example: single column
gap_1col <- gap %>% gs_read_cellfeed("Europe", range = cell_cols(3))
gap_1col

# convert to a un-named character vector and drop the variable name
gap_1col %>% gs_simplify_cellfeed(notation = "none", col_names = TRUE)
```

### Create sheets

You can use `googlesheets` to create new spreadsheets.

```{r new-sheet}
foo <- gs_new("foo")
foo
```

By default, there will be an empty worksheet called "Sheet1", but you can control it's title, extent, and initial data with additional arguments to `gs_new()` (see `gs_edit_cells()` in the next section). You can also add, rename, and delete worksheets within an existing sheet via `gs_ws_new()`, `gs_ws_rename()`, and `gs_ws_delete()`. Copy an entire spreadsheet with `gs_copy()`.

### Edit cells

You can modify the data in sheet cells via `gs_edit_cells()`. We'll work on the completely empty sheet created above, `foo`. If your edit populates the sheet with everything it should have, set `trim = TRUE` and we will resize the sheet to match the data. Then the nominal worksheet extent is much more informative (vs. the default of 1000 rows and 26 columns) and any future consumption via the cell feed will be much faster.

```{r edit-cells}
foo <- foo %>% gs_edit_cells(input = head(iris), trim = TRUE)
```

Go to [your Google Sheets home screen](https://docs.google.com/spreadsheets/u/0/), find the new sheet `foo` and look at it. You should see some iris data in the first (and only) worksheet. We'll also take a look at it here, by reading the data from `foo`.

Note how we always store the returned value from `gs_edit_cells()` (and all other sheet editing functions). That's because the registration info changes whenever we edit the sheet and we re-register it inside these functions, so this idiom will help you make sequential edits and queries to the same sheet.

```{r consume-edited-cells}
foo %>% gs_read()
```

Read the function documentation for `gs_edit_cells()` for how to specify where the data goes, via an anchor cell, and in which direction, via the shape of the input or the `byrow =` argument.

### Delete sheets

Let's clean up by deleting the `foo` spreadsheet we've been playing with.

```{r delete-sheet}
gs_delete(foo)
```

If you'd rather specify sheets for deletion by title, look at `gs_grepdel()` and `gs_vecdel()`. These functions also allow the deletion of multiple sheets at once.

### Upload delimited files or Excel workbooks

Here's how we can create a new spreadsheet from a suitable local file. First, we'll write then upload a comma-delimited excerpt from the iris data.

```{r new-sheet-from-file}
iris %>%
  head(5) %>%
  write.csv("iris.csv", row.names = FALSE)
iris_ss <- gs_upload("iris.csv")
iris_ss
iris_ss %>% gs_read()
file.remove("iris.csv")
```

Now we'll upload a multi-sheet Excel workbook. Slowly.

```{r new-sheet-from-xlsx}
gap_xlsx <- gs_upload(system.file("mini-gap.xlsx", package = "googlesheets"))
gap_xlsx
gap_xlsx %>% gs_read(ws = "Asia")
```

And we clean up after ourselves on Google Drive.

```{r delete-moar-sheets}
gs_vecdel(c("iris", "mini-gap"))
## achieves same as:
## gs_delete(iris_ss)
## gs_delete(gap_xlsx)
```

### Download sheets as csv, pdf, or xlsx file

You can download a Google Sheet as a csv, pdf, or xlsx file. Downloading the spreadsheet as a csv file will export the first worksheet (default) unless another worksheet is specified.

```{r export-sheet-as-csv}
gs_title("Gapminder") %>%
  gs_download(ws = "Africa", to = "~/tmp/gapminder-africa.csv")
## is it there? yes!
read.csv("~/tmp/gapminder-africa.csv") %>% head()
```

Download the entire spreadsheet as an Excel workbook.

```{r export-sheet-as-xlsx}
gs_title("Gapminder") %>% 
  gs_download(to = "~/tmp/gapminder.xlsx")
```

Go check it out in Excel, if you wish!

And now we clean up the downloaded files.

```{r clean-exported-files}
file.remove(file.path("~/tmp", c("gapminder.xlsx", "gapminder-africa.csv")))
```

### Authorization using OAuth2
 
If you use a function that requires authentication, it will be auto-triggered. But you can also initiate the process explicitly if you wish, like so:
 
```{r auth, eval = FALSE}
# Give googlesheets permission to access your spreadsheets and google drive
gs_auth() 
```
 
Use `gs_auth(new_user = TRUE)`, to force the process to begin anew. Otherwise, the credentials left behind will be used to refresh your access token as needed.

The function `gs_user()` will print and return some information about the current authenticated user and session.

```{r gs_user}
user_session_info <- gs_user()
user_session_info
```

### "Old" Google Sheets

In March 2014 [Google introduced "new" Sheets](https://support.google.com/docs/answer/3541068?hl=en). "New" Sheets and "old" sheets behave quite differently with respect to access via API and present a big headache for us. Recently, we've noted that Google is forcibly converting sheets: [all "old" Sheets will be switched over the "new" sheets during 2015](https://support.google.com/docs/answer/6082736?p=new_sheets_migrate&rd=1). However there are still "old" sheets lying around, so we've made some effort to support them, when it's easy to do so. But keep your expectations low.

In particular, `gs_read_csv()` does not currently work for "old" sheets.
