---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Project Status: Wip - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/0.1.0/wip.svg)](http://www.repostatus.org/#wip)[![Build Status](https://travis-ci.org/jennybc/gspreadr.svg?branch=master)](https://travis-ci.org/jennybc/gspreadr)[![Coverage Status](https://coveralls.io/repos/jennybc/gspreadr/badge.svg)](https://coveralls.io/r/jennybc/gspreadr)

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

---

Google Sheets R API
---

Manage your spreadsheets with *gspreadr* in R. 

*gspreadr* is inspired by [gspread](https://github.com/burnash/gspread), a Google Spreadsheets Python API

Features:

  - Access a spreadsheet by its title, key or URL.
  - Extract data or edit data.
  - Create | delete | rename | copy spreadsheets and worksheets.
  
### Install gspreadr

```{r eval = FALSE}
devtools::install_github("jennybc/gspreadr")
```

### Take a look at the vignette

This README is arguably as or more useful as the vignette and both are still under development. But feel free to [check out the current state of the vignette](http://htmlpreview.github.io/?https://raw.githubusercontent.com/jennybc/gspreadr/master/vignettes/basic-usage.html).

### Load gspreadr

`gspreadr` is designed for use with the `%>%` pipe operator and, to a lesser extent, the data-wrangling mentality of `dplyr`. But rest assured, neither is strictly necessary to use `gspreadr`. The examples here use both, but we'll soon develop a vignette that shows usage with plain vanilla R.

```{r load-package}
library("gspreadr")
suppressMessages(library("dplyr"))
```

### See some spreadsheets you can access

The `list_sheets()` function returns the sheets you would see in your Google Sheets home screen: <https://docs.google.com/spreadsheets/>. This should include sheets that you own and may also show sheets owned by others but that you are permitted to access, especially if you have clicked on a link shared by the owner. Expect a prompt to authenticate yourself in the browser at this point (more below re: authentication).

```{r list-sheets}
(my_sheets <- list_sheets())
# (expect a prompt to authenticate with Google interactively HERE)
my_sheets %>% glimpse()
```

### Register a spreadsheet

If you plan to consume data from a sheet or edit it, you must first register it. Basically this is where `gspreadr` makes a note of important info about the sheet that's needed to access via the Sheets API. Once registered, you can get some basic info about the sheet via `str()`.

```{r register-sheet}
# Hey let's look at the Gapminder data
gap <- register_ss("Gapminder")
str(gap)

# Need to access a sheet you do not own?
## Access it by key if you know it!
gap_key <- "1hS762lIJd2TRUTVOqoOP7g-h4MDQs6b2vhkTzohg8bE"
gap <- gap_key %>% register_ss

# gspreadr may be able to determine the key from the browser URL
gap_url <- "https://docs.google.com/spreadsheets/d/1hS762lIJd2TRUTVOqoOP7g-h4MDQs6b2vhkTzohg8bE/"
gap <- gap_url %>% register_ss
```

### Get a Google spreadsheet to practice with

If you don't have any suitable Google Sheets lying around, or if you just want to follow along verbatim with this vignette, this bit of code will copy a sheet from the `gspreadr` Google user into your Drive. The sheet holds some of the [Gapminder data](https://github.com/jennybc/gapminder).

```{r copy-gapminder, eval = FALSE}
gap_key <- "1hS762lIJd2TRUTVOqoOP7g-h4MDQs6b2vhkTzohg8bE"
copy_ss(key = gap_key, to = "Gapminder")
```

If that seems to have worked, go check that you see a sheet named Gapminder listed in your Google Sheets home screen: <https://docs.google.com/spreadsheets/>. You could also try `list_sheets()` again and make sure the Gapminder sheet is listed.

Now register your copy of the Gapminder sheet and you can follow along:

```{r register-gapminder, eval = FALSE}
gap <- register_ss("Gapminder")
str(gap)
```

### Consume data

There are two ways to consume data from a worksheet within a Google spreadsheet: the cell feed and the list feed. The cell feed gets data cell-by-cell. The list feed gets data by row. Read the function-level docs for more details about when to use which function.

```{r list-and-cell-feed}
# Get the data for worksheet "Oceania": the fast tabular way ("list feed")
oceania_list_feed <- gap %>% get_via_lf(ws = "Oceania") 
str(oceania_list_feed)
oceania_list_feed

# Get the data for worksheet "Oceania": the slower cell-by-cell way ("cell feed")
oceania_cell_feed <- gap %>% get_via_cf(ws = "Oceania") 
str(oceania_cell_feed)
head(oceania_cell_feed, 10)
```

#### Convenience wrappers and post-processing the data

There are a few ways to limit the data you're consuming. You can put direct limits into `get_via_cf()`, but there are also convenience functions to get a row (`get_row()`), a column (`get_col()`), or a range (`get_cells()`). Also, when you consume data via the cell feed (which these wrappers are doing under the hood), you will often want to reshape it or simplify it (`reshape_cf()` and `simplify_cf()`).

```{r wrappers-and-post-processing}
# Reshape: instead of one row per cell, make a nice rectangular data.frame
oceania_reshaped <- oceania_cell_feed %>% reshape_cf()
str(oceania_reshaped)
head(oceania_reshaped, 10)

# Limit data retrieval to certain cells

# Example: first 3 rows
gap_3rows <- gap %>% get_row("Europe", row = 1:3)
gap_3rows %>% head()

# convert to a data.frame (first row treated as header by default)
gap_3rows %>% reshape_cf()

# Example: first row only
gap_1row <- gap %>% get_row("Europe", row = 1)
gap_1row

# convert to a named character vector
gap_1row %>% simplify_cf()

# just 2 columns, converted to data.frame
gap %>%
  get_col("Oceania", col = 3:4) %>%
  reshape_cf()

# arbitrary cell range
gap %>%
  get_cells("Oceania", range = "D12:F15") %>%
  reshape_cf(header = FALSE)

# arbitrary cell range, alternative specification
gap %>%
  get_via_cf("Oceania", max_row = 5, min_col = 1, max_col = 3) %>%
  reshape_cf()
```

### Create sheets

You can use `gspreadr` to create new spreadsheets.

```{r new-sheet}
foo <- new_ss("foo")
foo %>% str
```

By default, there will be an empty worksheet called "Sheet1". You can also add, rename, and delete worksheets within an existing sheet via `add_ws()`, `rename_ws()`, and `delete_ws()`. Copy an entire spreadsheet with `copy_ss()`.

### Edit cells

You can modify the data in sheet cells via `edit_cells()`. We'll work on the completely empty sheet created above, `foo`.

```{r edit-cells}
foo <- foo %>% edit_cells(input = head(iris), header = TRUE)
```

Go to [your spreadsheets home page](https://docs.google.com/spreadsheets/u/0/), find the new sheet `foo` and look at it. You should see some iris data in the first (and only) worksheet. We'll also take a look at it here, by consuming `foo` via the list feed.

Note that we always store the returned value from `edit-cells()` (and all other sheet editing functions). That's because the registration info changes whenever we edit the sheet and we re-register it inside these functions, so this idiom will help you make sequential edits and queries to the same sheet.

```{r consume-edited-cells}
foo %>% get_via_lf() %>% print()
```

Read the function documentation for `edit_cells()` for ways to specify where the data goes and in which direction.

### Delete sheets

Let's clean up by deleting the `foo` spreadsheets we've been playing with.

```{r delete-sheet}
delete_ss("foo")
```

### Upload delimited files or Excel workbooks cells

Here's how we can create a new spreadsheet from a suitable local file. First, we'll write then upload a comma-delimited excerpt from the iris data.

```{r new-sheet-from-file}
iris %>% head(5) %>% write.csv("iris.csv", row.names = FALSE)
iris_ss <- upload_ss("iris.csv")
iris_ss %>% str()
iris_ss %>% get_via_lf() %>% print()
file.remove("iris.csv")
```

Now we'll upload a multi-sheet Excel workbook.

```{r new-sheet-from-xlsx}
upload_ss("tests/testthat/gap-data.xlsx")
gap_xlsx <- register_ss("gap-data")
gap_xlsx %>% str()
gap_xlsx %>% get_via_lf(ws = "Oceania") %>% print()
```

And we clean up after ourselves on Google Drive.

```{r delete-moar-sheets}
delete_ss("iris")
delete_ss("gap-data")
```

### Authorization using OAuth2
 
If you use a function that requires authentication, it will be auto-triggered. But you can also initiate the process explicitly if you wish, like so:
 
```{r authorize, eval = FALSE}
# Give gspreadr permission to access your spreadsheets and google drive
authorize() 
```
 
Use `authorize(new_user = TRUE)`, to force the process to begin anew. Otherwise, the credentials left behind will be used to refresh your access token as needed.

##### Stuff we are in the process of bringing back online after the Great Refactor of February 2015

  * visual overview of which cells are populated
