---
title: "OAuth Token Encryption for Travis"
output:
  html_document:
    keep_md: yes
    theme: cosmo
    toc: yes
    toc_depth: 4
---

*Heavily based on [Noam Ross's instructions](https://discuss.ropensci.org/t/test-api-wrapping-r-packages-with-oauth-tokens/157), here is how we obtain and encrypt the OAuth2.0 token used for testing, including on [Travis-CI](https://travis-ci.org). Go [here](http://docs.travis-ci.com/user/encrypting-files/) for the generic instructions on how to encrypt files on Travis.*

In theory, you do this once. In reality, you do it more often because something goes wrong with your token. For example, with Google, you can only have [25 active tokens per account](https://developers.google.com/identity/protocols/OAuth2#expiration). In the course of `googlesheets` development, if one of us does something that results in several new token requests, we inevitably use up some of those 25. Over time, the token on Travis eventually falls off the end and we need to request and encrypt a new one. Here's how we do it.

#### 0. Package design re: OAuth

The support provided by [`httr`](https://github.com/hadley/httr) for OAuth token management is a beautiful thing. Specifically, the automagic usage of `.httr-oauth` for token caching. It is a great kindness to preserve the user's blissful ignorance about the OAuth2 flow for as long as possible.

However, I've decided it's pretty important to anticipate token workflows where the user asserts more control. You will want this for development purposes, e.g., in automated testing against the API, and for non-interactive use of the package, e.g. in scripts run in batch mode or in a Shiny app. Our initial OAuth flow relied solely on `.httr-oauth` but we eventually extended `gs_auth()` to make it easy to retrieve the current token and to put a pre-existing token into force, whether stored in an R object or in an `.rds` file.

#### 1. Get a valid token and save it to file

Either do this from a working directory where the existing `.httr-oauth` contains the token you'd like to use or where there is NO existing `.httr-oauth` (or use `gs_auth(new_user = TRUE)`). In the latter case, when you are kicked to the browser, make sure you log in (or already logged in) to Google as the desired user. Pro tip: start with a fresh token or one near the beginning of the current 25-token sequence.

```{r eval = FALSE}
library(googlesheets)
token <- gs_auth()
saveRDS(token, file = "tests/testthat/googlesheets_token.rds")
```

This token is now in an obvious location for use in (local) testing via `testthat`.

#### 2. Use the stored token in your tests

Prior to tests that require authorization, put this token into force.

```{r eval = FALSE}
gs_auth(token = "googlesheets_token.rds")
```

or, more realistically,

```{r eval = FALSE}
suppressMessages(gs_auth(token = "googlesheets_token.rds", verbose = FALSE))
```

A line like this will be at the top of your testing files, but presumably outside of actual tests or expectations. Here's [an example](https://github.com/jennybc/googlesheets/blob/a1a4d877c7a0296f1c95e9f59b03ef4c1ac369b5/tests/testthat/test-gs-ls.R#L3). I also like to [explicitly suspend authenticated access](https://github.com/jennybc/googlesheets/blob/a1a4d877c7a0296f1c95e9f59b03ef4c1ac369b5/tests/testthat/test-gs-ls.R#L39) at the end of every testing file (note: `gs_auth_suspect()` is unexported ... hmmm, should I export so others could use?).

#### 3. Encrypt the token file and send to Travis

This is the work necessary to use our existing token on Travis CI, securely.

First you need to install the Travis command line client:

`gem install travis`

*Note: I had to do this as sudo.*

Then log into your Travis account using your GitHub username and password.

`travis login`

Encrypt the token and send to Travis:

`travis encrypt-file tests/testthat/googlesheets_token.rds --add`

The `--add` option should add a decrypt command to your pre-existing `.travis.yml` file, along these lines:

```
before_install:
- openssl aes-256-cbc -K $encrypted_84f43c85142c_key -iv $encrypted_84f43c85142c_iv -in tests/testthat/googlesheets_token.rds.enc -out tests/testthat/googlesheets_token.rds -d
```

*Double check that the token and encrypted token live in `tests/testthat/` and that `.travis.yml` reflects the correct path. You will probably need to move the encrypted token into the correct directory and edit the path(s) in `.travis.yml`!*

#### 4. Carefully ignore, commit, and push

List the token `tests/testthat/googlesheets_token.rds` in `.gitignore`. 

List the encrypted token `tests/testthat/googlesheets_token.rds.enc` in `.Rbuildignore`.

__Do not mix these up.__

  * Why do we gitignore the unencrypted token? You don't want your token file on GitHub.
  * Why do we NOT gitignore the encrypted token? Because then it would not go to GitHub and would not be available for decryption on Travis.
  * Why do we Rbuildignore the encrypted token? You don't want the encrypted version to be bundled and distributed with your package.
  * Why do we NOT Rbuildignore the unencrypted token? If you put `token_file.rds` in `.Rbuildignore`, it will not be copied over into the `my_package.Rcheck` directory when Travis runs `R CMD check`, and your tests will fail. Over and over. As you bang your desk trying to figure out what's wrong.
  * *Wait ... but doesn't this mean the unencrypted token __will__ go into the bundled package? Good point! See below for implications re: submission to win-builder and CRAN.*

Commit `tests/testthat/googlesheets_token.rds.enc` and your updated `.travis.yml` and push to Github. God willing, your tests that require authorization will now pass on Travis.

#### 5. Special considerations when submitting to win-builder and CRAN

I use the token in `tests/testthat/googlesheets_token.rds` to authorize `googlesheets` in the vignette and was sort of horrified when I noticed that the vignette compiled *just fine* on <http://win-builder.r-project.org>. Since we do NOT (and cannot) Rbuildignore the unencrypted token, it was indeed being bundled up and sent along to win-builder. This got me concerned that, unless I took deliberate steps to omit the token in my eventual CRAN submission, it would be distributed with my package source.

I decided to make a "CRAN submission" branch and add `tests/testthat/googlesheets_token.rds` to `.Rbuildignore`. Of course, this breaks the build on Travis, so this little exercise will have to be done just prior to each CRAN submission and I'll need to take care to submit from the branch, not from master.

When I first submitted the "CRAN submission" branch to win-builder, I got an education on all the ways in which CRAN checks vignettes. In the absence of the token, `googlesheets` cannot do anything that requires authorization, which broke many chunks of the original vignette. How I worked around that is a separate story *(TO DO: write that down)*.

Temporary summary of vignette analogue of `testthat::skip_on_CRAN()`: there are two issues. First, CRAN will attempt to build the vignette. Second, CRAN will attempt to run the R code extracted from the vignette. To solve the first problem, conditionally suppress chunk execution on CRAN. To solve the second problem, prevent the extraction of source code from some or all chunks with `purl = FALSE`. The vignette CRAN actually uses is the one you build and submit, so these measures aren't as harmful as you might fear.

For CRAN updates, I still plan to vet them with win-builder and it's pretty clear that, if I allow the token to go, it's not kept private in any shape or form. So either I'll just tolerate the small chance this could come back to haunt me or I'll handle like a CRAN submission and submit from a branch in which I've Rbuildignored the unencrypted token.

Notes
---

2015-06-29 The token we had been using finally fell off the end of the 25-token stack. But the basic instructions did not work for me when attempting to encrypt the new token file. I got this:

```
Jennifers-2015-MacBook-Pro:googlesheets jenny$ travis encrypt-file tests/testthat/googlesheets_token.rds 
repository not known to https://api.travis-ci.com/: jennybc/gspreadr
```

The old name of this repo was somehow blocking me. The repo hasn't been called `gspreadr` for months! From [this issue thread](https://github.com/travis-ci/travis-ci/issues/3093) I learned to inspect and correct the Travis slug in `.git/config`. It now reads like so:

```
[travis]
	slug = jennybc/googlesheets
```

That allowed me to encrypt a new token file.
