---
title: "My Google Sheets Explorer"
output: html_document
---

```{r}
library(googlesheets)
```

## Description

This app is currently deployed [here](https://jozhao.shinyapps.io/gs-explorer/). 

**My Google Sheets Explorer** allows any user to log in using their Google account. After authorization, the user is presented with a listing of all their Google Sheets. The user is allowed to select a Sheet of interest and additional details about the Sheet will be presented in the `Sheet Info` tab. The user is also allowed to quickly inspect the worksheets that live in the selected Sheet by going to the `Sheet Inspection` tab. Simply select a worksheet and a visual representation of the worksheet is displayed.

---

## Running app locally 

`googlesheets` currently has set up web application credentials in Google Developer Console. If you are creating an app locally, default app credentials are built into `googlesheets`, just run your app using `runApp(port = 4642)`. Credentials are accessible through:

```{r}
getOption("googlesheets.shiny.client_id")
getOption("googlesheets.shiny.client_secret")
getOption("googlesheets.shiny.redirect_uri")
```

Two functions are used to get authorization working: `gs_shiny_get_url()` and `gs_shiny_get_token()`.

The general flow is:

1. User clicks url formed by `gs_shiny_get_url()`.  
```{r}
# looks like this
gs_shiny_get_url()
```
2. User authenticates is redirected back to app. An authorization code also gets returned in the url in which it can be extracted using `parseQueryString(session$clientData$url_search)$code`. 

3. Extract the authorization code and make a POST request to exchange for an access token using `gs_shiny_get_token()`. 

## Setting up authorization in your own app

You must create your own web application in [Google Developer's Console](https://console.developers.google.com/). Obtain the client ID, client secret and set the redirect URI to some static local url ie. `http://127.0.0.1:4642` (for testing so every time you would run `runApp(port = 4642)` ). Depending on where you will deploy your app, you will later come back and add that destination as another redirect uri. 

After setting up the web application, in your `server.R` set client ID and secret, and redirect uri specific to your project by:

```
options("googlesheets.shiny.client_id" = MY_CLIENT_ID)
options("googlesheets.shiny.client_secret" = MY_CLIENT_SECRET)
options("googlesheets.shiny.redirect_uri" = MY_REDIRECT_URI)
```



