---
title: "gspreadr Basic Usage"
author: "Joanna Zhao"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette gspeadr Basic Usage}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r load package}
library(gspreadr)
```

This vignette shows the basic functionality of `gspreadr`.

# User Authentication

In order to access your own spreadsheets (non-public), you must log in. There are 2 ways to do it in `gspreadr`. 

1.  `login("email", "password")` (if you really want to...)
2.  `authorize()` (recommended)

For the first time, `authorize()` will authenticate the user interactively. The user will be directed to a web browser where they will be asked to login to their Google account, and give `gspreadr` permission to access Sheets and Google Drive. Credentials will be cached and authentication will not be needed for subsequent requests/sessions as `httr` handles the storing and refreshing of tokens. (A `.httr-oauth` file will be left behind in the current working directory)

If you want to access a different account, run `authorize(new_user = TRUE)`, as this will delete the previously stored token and get a new one for the new account.

# Open a spreadsheet

## For public spreadsheets

To access public spreadsheets, you will either need the key of the spreadsheet (as found in the URL) or the entire URL. 

```{r open public sheets, eval = FALSE}

pub_sheet <- open_by_key("1hS762lIJd2TRUTVOqoOP7g-h4MDQs6b2vhkTzohg8bE", visibility = "public")

pub_sheet <- open_by_url("https://docs.google.com/spreadsheets/d/1hS762lIJd2TRUTVOqoOP7g-h4MDQs6b2vhkTzohg8bE/pubhtml", visibility = "public")

```

## For private spreadsheets

Once you have authenticated, you can use `list_spreadsheets()` to remind yourself what spreadsheets are in your Google drive. A spreadsheet can be opened by its title.

```{r spreadsheet ops, eval = FALSE}
# list spreadsheet titles, owners, and time of last update
list_spreadsheets()

# list the spreadsheet key too
list_spreadsheets(show_key = TRUE)

ssheet <- open_spreadsheet("Gapminder")
```

```{r, include = FALSE }
ssheet <- open_by_url("https://docs.google.com/spreadsheets/d/1hS762lIJd2TRUTVOqoOP7g-h4MDQs6b2vhkTzohg8bE", visibility = "public")
```

This returns a spreadsheet object. You can get the spreadsheet id, title, time of last update, and the number and names of worksheets contained. 

```{r}
# spreadsheet components
ssheet$sheet_id ; ssheet$sheet_title ; ssheet$updated ; ssheet$nsheets ; ssheet$ws_names
```

## Add or delete a spreadsheet

To add or delete a spreadsheet in your Google Drive, use `add_spreadsheet()` or `del_spreadsheet()` and simply pass in the title of the spreadsheet as a character string. The new spreadsheet by default will contain one worksheet titled "Sheet1".

```{r create and delete spreadsheet, eval = FALSE}
# Create a new spreadsheet by title
add_spreadsheet("New Spreadsheet")

# Move spreadsheet to trash
del_spreadsheet("New Spreadsheet")
```

# Open a worksheet

You can use `list_worksheets()` to see the worksheets living in a spreadsheet. This is a wrapper around `ssheet$ws_names`.
```{r list ws}

list_worksheets(ssheet)
```

To open a worksheet, you can pass in a spreadsheet object and then specify the worksheet using its title or index, or this can be done in one step using `open_at_once(ss_title, ws_index)`. This saves a bit of typing but the spreadsheet object will not be stored, so if you want to open another worksheet, `open_spreadsheet("title")` will be called again. So if you want to open multiple worksheets from a spreadsheet, it is best to open the spreadsheet and save the spreadsheet as an R object and call `open_worksheet(ss, x)`  

```{r open ws}
# by title
ws <- open_worksheet(ssheet, "Oceania")
```
```{r, eval=FALSE}
# or by index (starting at 1)
ws <- open_worksheet(ssheet, 5)

# or in one shot
ws <- open_at_once(ss_title = "Gapminder by Continent", ws_index = "Oceania")
ws <- open_at_once(ss_title = "Gapminder by Continent", ws_index = 5)
```

This returns a worksheet object. You can get the ids of the spreadsheet and worksheet, worksheet title, and the number of rows and columns.

```{r, eval = FALSE}
# worksheet components
ws$sheet_id ; ws$id ; ws$title ; ws$ncol ; ws$nrow
```

# Open multiple worksheets

You can open all the worksheets from a spreadsheet using `open_worksheets()` passing the spreadsheet object as the argument. This will return a list of worksheet objects which you can then use `plyr` functions to perform worksheet operations such as `get_cell()`, `get_row()`, `read_region()`, etc.   

```{r open multiple worksheets, eval = FALSE}

ws_objs <- open_worksheets(ssheet)

plyr::llply(ws_objs, function(x) get_cell(x, "A2"))

```

# Add or delete a worksheet

To add a worksheet to a spreadsheet, pass in the spreadsheet object, title of new worksheet and the number of rows and columns. To delete a worksheet from a spreadsheet, pass in the spreadsheet object and the title of the worksheet. Note that after adding or deleting a worksheet, the spreadsheet object will not be automatically updated to include the new worksheet(s) information, you must open the spreadsheet again to "refresh" it. 

```{r, eval = FALSE}
add_worksheet(ssheet, title = "foo", rows = 10, cols = 10)

del_worksheet(ssheet, ws_title)
```

# Rename a worksheet

To rename a worksheet, pass in the spreadsheet object, the worksheet's current name and the new name you want it to be.  

```{r eval = FALSE}
rename_worksheet(ssheet, "Sheet1", "First Sheet")
```

# Worksheet Operations

## View worksheet

You can take a look at your worksheets to get an idea of what it looks like. Use `view()` to look at one worksheet and `view_all()` to look at all worksheets contained in a spreadsheet. `view_all()` returns a gallery of all the worksheets. Set `show_overlay = TRUE` to view an overlay of all the worksheets to identify the density of the cells occupied by worksheets. **showing Error: could not find function "ggplotGrob"**

```{r, fig.width=7, fig.height=7}

view(ws)

view_all(ssheet)
```

## Display structure of spreadsheet and worksheet

To get a more detailed report of your worksheet, use `str()`. This function summarizes your worksheet by letting you know the number of rows and columns, empty cells, and the cell pattern per column (a string of ##V for values and ##NA for empty cells).
```{r display structure}

# str for a spreadsheet
str(ssheet)

# str for a worksheet
str(ws)

```

## Get data from a worksheet

You can get row(s), col(s), a region, or an entire worksheet. 

### Get row(s)
```{r get rows}
get_row(ws, 1)

get_rows(ws, from = 1, to = 3)
```

### Get col(s)

You can get columns either by integer or letter. 

```{r get cols}
head(get_col(ws, 1))

head(get_col(ws, "A"))

head(get_cols(ws, from = 1, to = 5))

head(get_cols(ws, from = "A", to = "E"))

```

### Get a region as a data frame

#### Specify by min/max rows and columns

You can get a region of a worksheet by specifying the boundary rows and columns. The first row will be taken as the `header` by default.

```{r get region}
read_region(ws, from_row = 1, to_row = 3, from_col = 1, to_col = 5)

# set header to FALSE
read_region(ws, from_row = 1, to_row = 3, from_col = 1, to_col = 5, header = FALSE)
```

#### Specify by range

You can also get a region of a worksheet by specifying the cell ranges as a character string. The first row will be taken as the `header` by default.

```{r get range}

read_range(ws, "A1:E5")

```


### Get the entire worksheet as data frame

You can get the entire worksheet. Empty cells will be filled with NAs. Boundaries are determined by the right-most and bottom-most cells that contain a value.  The first row will be taken as the `header` by default.


```{r get all, eval = FALSE}
read_all(ws)
```

### Get a cell

You can get the value of a specific cell. There are two notations you can use: `A1` and `R1C1`. For `A1` notation, columns are indicated by a letter followed by the row as an integer. For `R1C1` notation, integers represent the rows and columns, and are preceded by `R` or `C`. Note that `A1` and `R1C1` are equivalent.

```{r get cell}

get_cell(ws, "A1")

get_cell(ws, "R1C1")

```

### Find a cell

You can find the first position of the cell containing the specified value or all the cells that contain the specified value. 

```{r find cell}

find_cell(ws, "Australia")

find_all(ws, "Australia")

```

## Update cells

### Update a single cell

You can update a cell's value by specifying the cell's position, either in `A1` or `R1C1` notation, and the new value. The new value may be a formula followed by a `=`, ie. `=A1+B1`. 

```{r update cell, eval = FALSE}

update_cell(ws, "A1", "Oops")

get_cell(ws, "A1")

update_cell(ws, "R1C1", "country")

get_cell(ws, "R1C1")

```

### Update cells in batch

You can update a batch of cells by specifying the range (ie. "A1:A4"). Alternatively, if you just want to dump an entire dataframe or vector into a worksheet, you can specify an anchor cell as the reference cell position and the range will be calculated for you. You can pass in a vector of new values or an entire data frame.

```{r update cells, eval = FALSE}

update_cells(ws, "C1:E1", c("A", "B", "C"))

read_range(ws, "A1:F3")

update_cells(ws, "G1", head(iris))

read_range(ws, "G1:K7")
```

```{r return to orig, include = FALSE, eval = FALSE}

update_cells(ws, "C1:E1", c("pop", "continent", "lifeExp"))
update_cells(ws, "G1:K7", ncol(head(iris) * (nrow(iris) + 1)))
```

