---
title: "gspreadr Basic Usage"
author: "Joanna Zhao"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette gspeadr Basic Usage}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r load package}
library(gspreadr)
```

This vignette shows the basic functionality of `gspreadr`.

## User Authentication

In order to access your own spreadsheets (non-public), you must log in. There are 2 ways to do it in `gspreadr`. 

1.  `login("email", "password")` (if you really want to...)
2.  `authorize()` (preferred)

For the first time, `authorize()` will authenticate the user interactively. The user will be directed to a web browser where they will be asked to login to their Google account, and give `gspreadr` permission to access Sheets and Google Drive. Credentials will be cached and authentication will not be needed for subsequent requests/sessions as `httr` handles the storing and refreshing of tokens.

If you want to access a different account, run `authorize(new_user = TRUE)`, as this will delete the previously stored token and get a new one for the new account.

## Open a spreadsheet

You can use `list_spreadsheets()` to remind yourself what spreadsheets are in your Google drive. A spreadsheet can be opened by its title.

```{r spreadsheet ops}
list_spreadsheets()

ssheet <- open_spreadsheet("basic-usage")
```
This returns a spreadsheet object. You can get the spreadsheet id, title, time of last update, and the number and names of worksheets contained. 

```{r}
# spreadsheet attributes
ssheet$sheet_id ; ssheet$sheet_title ; ssheet$updated ; ssheet$nsheets ; ssheet$ws_names
```

## Open a worksheet

You can use `list_worksheets()` to see the worksheets living in the spreadsheet. This is a wrapper around `ssheet$ws_names`.
```{r list ws}

list_worksheets(ssheet)
```

To open a worksheet, you can pass in a spreadsheet object and then specify the worksheet using its title or index, or this can be done in one step using `open_at_once(ss_title, ws_index)`. This saves a bit of typing but the spreadsheet object will not be stored, so if you want to open another worksheet, `open_spreadsheet("title")` will be called again. So if you want to open multiple worksheets from a spreadsheet, save the spreadsheet as an R object and use it for `open_worksheet(ss, x)`  

```{r open ws}

# by title
ws <- open_worksheet(ssheet, "Sheet1")
```
```{r, eval=FALSE}
# or by index
ws <- open_worksheet(ssheet, 1)

# or in one shot
ws <- open_at_once(ss_title = "basic-usage", ws_index = "Sheet1")
ws <- open_at_once(ss_title = "basic-usage", ws_index = 1)
```

This returns a worksheet object. You can get the ids of the spreadsheet and worksheet, worksheet title, and the number of rows and columns.

```{r}
# worksheet attributes
ws$sheet_id ; ws$id ; ws$rows ; ws$cols
```

## Operations on a worksheet

You can get row(s), col(s), a region, or an entire worksheet. 

### Get row(s)
```{r get rows}
get_row(ws, 1)

get_rows(ws, from = 1, to = 3)
```

### Get col(s)
```{r get cols}
get_col(ws, 1)

get_cols(ws, from = 1, to = 5)
```

### Get a region as a data frame

You can get a region of a worksheet by specifying the boundary rows and columns. The first row will be taken as the `header` by default.

```{r get region}
get_region(ws, from_row = 1, to_row = 6, from_col = 1, to_col = 5)

# set header to FALSE
get_region(ws, from_row = 1, to_row = 6, from_col = 1, to_col = 5, header = FALSE)
```
### Get the entire worksheet as data frame

You can get the entire worksheet. Empty cells will be filled with NAs. Boundaries are determined by the right-most and bottom-most cells that contain a value.  The first row will be taken as the `header` by default.


```{r get all}
get_all(ws)
```

### Add a worksheet
```{r, eval = FALSE}
add_worksheet(ssheet, title = "foo", rows = 10, cols = 10)
```

### Delete a worksheet
```{r, eval = FALSE}
del_worksheet(ssheet, ws)
```



